Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-02-23 analysis/paper/paper.Rmd
changed map file name
Incorporated edits from coauthors

---
title: "Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations"
author:
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [mhe_mccall, biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Fish Program, Science Division, 1111 Washington Street NE, Olympia, Washington, 98501, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA 
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
keywords: |
  Northern Pikeminnow; Chinook Salmon; predation; mark-recapture; bioenergetics
---

@@

Predation of emigrating juvenile salmonids by piscivorous fishes is perhaps an under-appreciated source of mortality hindering population recovery efforts in the Columbia River basin. Reservoirs and tailraces associated with dams on the Snake and Columbia river have created favorable conditions for fishes known to consume juvenile salmonids. The native Northern Pikeminnow are one of these fishes and have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above these reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider potential impacts of them on local Chinook Salmon populations. The reach was (at least partially) formed by the failure of a small mining dam in a nearby drainage during the late 19th century and contains historically important overwinter rearing habitat and is a migration corridor for Chinook Salmon as well as other Endangered Species Act-listed species including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough to be 27,874 in the fall and 19,499 in the spring during the peak emigrations of juvenile Chinook Salmon. Assuming a scenario where 60% of the Northern Pikeminnow diet is fish, of which 50% is Chinook Salmon, we estimated the Northern Pikeminnow population to consume 95,120 juvenile Chinook Salmon, annually. In the most conservative scenario where 30% of diet is fish and 30% of fish are juvenile Chinook Salmon, 47,079 juveniles are consumed. Using a sensitivity analysis, we further considered potential impacts to adult returns to local populations. Given the size of the Salmon River, the estimated Northern Pikeminnow densities for Deadwater Slough are remarkable and predation from the population likely has consequential impacts on Chinook Salmon in the Upper Salmon River major population group, thereby hindering local recovery efforts.

@@

One potentially important, but perhaps under-appreciated source of salmonid mortality is predation on emigrating juveniles by piscivorous fishes. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of recovery [@Fresh2003; @Winther2020]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Indeed, Northern Pikeminnow are estimated to consume ~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is in a river reach containing historically important overwinter rearing habitat, includes the emigration corridor for juvenile Chinook Salmon *O. tshawytscha*, and is inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*. Further, the slough is part of the migratory pathway of the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead.

Quantifying mortality associated with piscine predation in upstream habitats of the Salmon River is of particular interest for spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon major population group (MPG). The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Currently, the Upper Salmon MPG supports eight independent, extant populations; this includes Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River populations [@NOAA2017]. The Lemhi River historically contained the largest Chinook Salmon population within the MPG and is therefore of particular interest for restoration actions [@Zimmerman2012]. The Upper Salmon MPG are stream-type Chinook Salmon and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency [@Dodson2013]. Deadwater Slough represents an important habitat for the Upper Salmon MPG as is supports rearing and migration of juveniles from all populations including fall DSR and spring NRR migrants.

@@

1. Estimate the abundance of Northern Pikeminnow in the Deadwater Slough during the peaks of fall DSR and spring NRR juvenile emigrations;
2. Document predation on juvenile Chinook Salmon during the DSR and NRR emigration periods using gastric lavage;
3. Use a bioenergetics approach to estimate consumption potential (grams) of the Northern Pikeminnow population on DSR and NRR Chinook Salmon emigrants at Deadwater Slough;
4. Estimate how consumption of juvenile Chinook Salmon emigrants at Deadwater Slough by Northern Pikeminnow may impact adult returns to the Upper Salmon MPG.

@@

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located roughly 6 river kilometers (rkm) downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event and upslope instability that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973]. That sediment formed an alluvial fan which created a slow, deep section in the river, spanning approximately `r round(ddw_hectares)` hectares, averaging `r round(ddw_avg_width_m)` m wide, and resembling a small reservoir.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. The initial intent was to survey during the emigration of the 2018 brood year (fall 2019 and spring 2020), however logistical constraints during the onset of the Covid-19 pandemic delayed the spring 2020 survey until fall 2020. Although this allowed an additional survey during the fall DSR emigration, surveys during the spring NRR emigration were still lacking. Consequently, an additional week of surveying was added in spring 2021. The fall surveys were constrained to two weeks to minimize Northern Pikeminnow immigration/emigration.

Multiple capture methods were employed during the initial fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to ESA-listed adult steelhead that were present in the Salmon River during our surveys. Ineffective methods were abandoned in subsequent years and the following analyses will focus on fish captured by angling, unless otherwise noted.

@@

Two types of mark-recapture estimators were used to explore variation in abundance estimates: a single census and a multiple census estimator. For the single census estimator, we treated the first week of the survey as the mark event and the following week as the recapture event, pooling data within each week. Alternatively, the multiple census estimator treated each day as an event and used information about the total number of marked fish from all previous events to calculate the total abundance.

For the single census estimator, we used the Chapman-modified Lincoln-Peterson estimator:

@@

where $M$ is the total number of fish marked and returned to the population during the first event (week), $n$ is the total number of fish caught in the second event (week), and $m$ is the number of marked fish caught during the second week. For a multiple census estimator, we used the Schnabel method:

@@

Because captures of new and marked fish were dependent on volunteer anglers, there was concern that sampling was biased to locations preferred by anglers. If marked fish were not recently released or mixed back into those preferred locales, marked fish may have been less susceptible to capture than unmarked fish. To account for this potential bias, we also used an adjusted Schnabel estimator, which assumed that marked and released fish during each survey day were not available for recapture for 48 hours. The adjusted Schnabel estimator, referred to as the "delayed mixing" Schnabel estimator, is given by:

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_{i-1}}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and 2020 surveys. However, to estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of total CPUE in the fall survey to the total CPUE in the spring survey, then multiplied that ratio by the mean abundance from fall surveys.

@@

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into linear and areal densities by dividing by the length and area, respectively, of Deadwater Slough measured in GIS software and using drone-generated orthography.

All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the survey period), (2) all fish have equal chance of being caught in subsequent survey days, (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or mistaken.

@@

Using hook-and-line angling, we caught `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "n_fish"])`, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "n_fish"])`, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "n_fish"])` Northern Pikeminnow during the fall 2019, fall 2020, and spring 2021 surveys, respectively, for a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` fish. The CPUE was `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "catch_per_hr"], digits = 3)` for fall 2019, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "catch_per_hr"], digits = 3)` for fall 2020, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "catch_per_hr"], digits = 2)` for spring 2021. During the mark-recapture events, there were `r as.character(english(sum(N_mods[[3]][[1]]$m)))` recaptures in fall 2019, including `r as.character(english(LP_mod[[3]][[1]]$m))` unique individuals and `r as.character(english(sum(N_mods[[3]][[3]]$m)))` recaptures in fall 2020, including `r as.character(english(LP_mod[[3]][[3]]$m))` unique individuals. Summaries of parameters for both the single census and multiple census mark-recapture estimators are provided in Table \@ref(tab:abund-data-lp) and Table \@ref(tab:abund-data-schn), respectively.

Mark-recapture abundance estimates for Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` depending on the estimator used (Table \@ref(tab:abund-tab)). For both surveys, the multiple census estimates were generally larger than the single census estimates and small numbers of recaptures resulted in wide confidence intervals (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`). Using the delayed-mixing Schnabel estimator, the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

The lengths of Northern Pikeminnow captured ranged from `r min(cmr_df$length, na.rm = T)` to `r max(cmr_df$length, na.rm = T)` mm TL with an average of `r round(mean(cmr_df$length, na.rm = T))` mm (Figure \@ref(fig:length-freq-p)). The PSD for Northern Pikeminnow in Deadwater Slough across all three efforts was `r prettyNum(psd, digits = 1)`%.

@@

In addition to the large population, the observed PSD (`r round(psd)`%) in this study was large compared to observations of 41% below Bonneville Dam and 18% in Bonneville Reservoir [@Winther2020]. This suggests that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. Notably, the state catch-and-release record Northern Pikeminnow, measuring `r max(gastric_df$length, na.rm = T)` mm TL, was caught in Deadwater Slough during the fall 2020 survey.

@@

```{r map, out.width = "85%", fig.cap = "Map of the Deadwater Slough study area. The high-resolution orthoimage portion surrounding the Deadwater Slough was generated from aerial images taken by an unmanned aircraft system (drone). The red polygon indicates the reach characterized by unnaturally slow water velocities and a deepened channel. The location of the Dump Creek delta is indicated."}
knitr::include_graphics("../figures/deadwater_map_low_res.png")

>>>>

---
title: "Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations"
author:
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [mhe_mccall, biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Fish Program, Science Division, 1111 Washington Street NE, Olympia, Washington, 98501, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA 
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
keywords: |
  Northern Pikeminnow; Chinook Salmon; predation; mark-recapture; bioenergetics
---

@@

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. The native Northern Pikeminnow are one of these predators and have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider potential impacts of them on local Chinook Salmon populations. The Deadwater Slough contains historically important overwinter rearing habitat and is a migration corridor for Chinook Salmon as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough during the peak emigrations of juvenile Chinook Salmonto be 27,874 in the fall and 19,499 in the spring. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon for varying proportions of fish in their diet, ranging from 30% to 90%. Assuming a median scenario where 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 95,120 juvenile Chinook Salmon, annually. Using a sensitivity analysis, the potential impact of predation on juvenile Chinook Salmon under the median scenario is a reduction in adult returns by 539 (95% CI: 231 - 1,335). Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impacts on Chinook Salmon in the Upper Salmon River major population group.

@@

Predation on emigrating juveniles by piscivorous fishes is another important, and potentially under-estimated, source of salmonid mortality. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of recovery [@Fresh2003; @Winther2020]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow-water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Northern Pikeminnow are estimated to consume ~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough contains historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu* that are potential sources of mortality for these rearing and migrating salmonids.

Quantifying mortality associated with piscine predation in upstream habitats of the Salmon River is of particular interest for spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon major population group (MPG). The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Currently, the Upper Salmon MPG supports eight independent, extant populations; this includes Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River populations [@NOAA2017]. The Lemhi River historically contained the largest Chinook Salmon population within the MPG and is therefore a prioritized candidate for restoration of natural processes [@Zimmerman2012].

The Upper Salmon MPG are stream-type Chinook Salmon and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency [@Dodson2013]. Deadwater Slough represents an important habitat for the Upper Salmon MPG as it supports rearing and migration of juveniles from all populations including fall DSR and spring NRR migrants.

@@

1. Estimate the abundance of Northern Pikeminnow in the Deadwater Slough during the peaks of fall DSR and spring NRR juvenile emigrations;
2. Document predation on juvenile Chinook Salmon during the DSR and NRR emigration periods using gastric lavage;
3. Use an established bioenergetics approach to estimate consumption potential (grams) of the Northern Pikeminnow population on DSR and NRR Chinook Salmon emigrants at Deadwater Slough;
4. Estimate how consumption of juvenile Chinook Salmon emigrants at Deadwater Slough by Northern Pikeminnow may impact adult returns to the Upper Salmon MPG.

@@

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located roughly 6 river kilometers (rkm) downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event and upslope instability that deposited substantial amounts of sediment at its confluence with the Salmon River [@Emerson1973]. That sediment formed an alluvial fan which created a slow, deep section in the river, spanning approximately `r round(ddw_hectares)` hectares, averaging `r round(ddw_avg_width_m)` m wide, and resembling a small reservoir.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. The initial intent was to survey during the emigration of the 2018 Chinook Salmon brood year (fall 2019 and spring 2020), however logistical constraints during the onset of the Covid-19 pandemic delayed the spring 2020 survey until 2021. Over this period, an additional fall survey occurred during the 2020 DSR emigration. The fall surveys were constrained to two weeks to minimize Northern Pikeminnow immigration/emigration.

Multiple capture methods were employed during the initial fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to ESA-listed adult steelhead that were present during our surveys. Ineffective methods were abandoned in subsequent years and the following analyses will focus on fish captured by angling, unless otherwise noted.

@@

Two types of mark-recapture estimators were used to explore variation in abundance estimates: a single census and a multiple census estimator. For the single census estimator, we treated the first week of the survey as the mark event and the following week as the recapture event, pooling data within each week. Alternatively, the multiple census estimator treated each day as a sampling event and used information about the total number of marked fish from all previous events to calculate the total abundance. For the single census estimator, we used the Chapman-modified Lincoln-Peterson estimator:

@@

where $M$ is the total number of fish marked and returned to the population during the first event (week), $n$ is the total number of fish caught in the second event (week), and $m$ is the number of marked fish caught during the second week. For a multiple census estimator, we used the Chapman-modified Schnabel estimator:

@@

Because captures of new and marked fish were dependent on volunteer anglers, there was concern that sampling was biased to locations preferred by anglers. If marked fish were not recently released or mixed back into those preferred locales, marked fish may have been less susceptible to capture than unmarked fish. To account for this potential bias, we also used an adjusted Schnabel estimator, which assumed that marked and released fish during each survey day were not available for recapture for 48 hours. The adjusted Schnabel estimator, referred to as the "Delayed-mixing" Schnabel estimator, is given by:

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_{i-1}}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$
All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the survey period), (2) all fish have equal chance of being caught in subsequent survey days, (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or mistaken.

To estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of mean total CPUE in the fall surveys to the total CPUE in the spring survey, then multiplied that ratio by the mean abundance from fall surveys.

@@

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into linear and areal densities by dividing by the length and area of Deadwater Slough. Length and area were measured in GIS software and using drone-generated orthoimagery.

@@

Using hook-and-line angling, we caught `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "n_fish"])`, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "n_fish"])`, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "n_fish"])` Northern Pikeminnow during the fall 2019, fall 2020, and spring 2021 surveys, respectively, for a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` fish. The CPUE was `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "catch_per_hr"], digits = 3)` for fall 2019, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "catch_per_hr"], digits = 3)` for fall 2020, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "catch_per_hr"], digits = 2)` for spring 2021. During the mark-recapture events, there were `r as.character(english(sum(N_mods[[3]][[1]]$m)))` recaptures in fall 2019, including `r as.character(english(LP_mod[[3]][[1]]$m))` unique individuals and `r as.character(english(sum(N_mods[[3]][[3]]$m)))` recaptures in fall 2020, including `r as.character(english(LP_mod[[3]][[3]]$m))` unique individuals. Summaries of parameters for both the single census and multiple census mark-recapture estimators are provided in Table \@ref(tab:abund-data-lp) and Table \@ref(tab:abund-data-schn).

Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` depending on the estimator used (Table \@ref(tab:abund-tab)). For both surveys, the unadjusted multiple-census estimates were larger than the single census estimates and small numbers of recaptures resulted in wide confidence intervals (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`). Using the delayed-mixing Schnabel estimator, the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

The lengths of Northern Pikeminnow ranged from `r min(cmr_df$length, na.rm = T)` to `r max(cmr_df$length, na.rm = T)` mm TL with an average of `r round(mean(cmr_df$length, na.rm = T))` mm (Figure \@ref(fig:length-freq-p)). The PSD for Northern Pikeminnow in Deadwater Slough across all three surveys was `r prettyNum(psd, digits = 1)`%.

@@

In addition to the large population, the observed (`r round(psd)`%) PSD in this study was large compared to observations of 41% below Bonneville Dam and 18% in Bonneville Reservoir [@Winther2020]. This suggests that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. Notably, the Idaho state catch-and-release record Northern Pikeminnow, measuring `r max(gastric_df$length, na.rm = T)` mm TL, was caught in Deadwater Slough during the fall 2020 survey.

@@

```{r map, out.width = "85%", fig.cap = "Map of the Deadwater Slough study area within the Upper Salmon River MPG. The Dump Creek alluvial fan is loacted at the downstream end of the study reach."}
knitr::include_graphics("../figures/Study Area Map 5.png")
